On Devon Allen's Disqualification at the 2022 World Track and Field Championships

### 2023-08-03 Manuscript/manuscript.tex
Overhealued GLMM results

\jy{Use double blank lines to separate paragraphs in tex source. Acknowledge if you copy.}

@@

The dataset of the four United States athletes 
is too small to perform any meaningful statistical analysis, but
the data was expanded to include reaction times for athletes who competed at 
national competitions (many of which were held from May-July of 2022). Once the 
data was collected, each athlete was designated as a cluster and the stage of
competition (national or international) was specified for each competition 
within the cluster (athlete).  We chose to include preliminary rounds in this
section as well as in \ref{sec:data2019} in order to increase the size of our
data.  Without data from the preliminary rounds, the number of observations
dropped from 65 to 26 and the number of athletes included in the analysis
decreased from 15 to 9.  Likewise

@@

\of{Add something like this to the start of the paragraph?}
We may want to consider analyzing all four data sets separately because
comparing the model based on all the data to the model based on data excluding 
2022 can provide an indication about the impact of the 2022 data.
We hope to see that by including 2022 data,
the probability of observing a significantly low reaction time increases. 
Table~\ref{tab:Gamma_parameters} summarizes the results from the gamma model 
from different data sets and shows the standard error of the year
effect increases for both the men's final and men's pooled data sets when
including 2022.  The men's final data venue effect error increased from 0.439
to 0.504 and from 0.0495 to 0.0554 for the men's pooled data.  This demonstrates
that the inclusion of 2022 increased the randomness of the data and motivates
the rank-based study (see Section~\ref{subsec:Results_Rank}).

\begin{table}
  \centering
  \caption{Summary statistics for the GLMM on various data sets.} 
  \begin{tabular}{c c c c c c}
      \toprule
      Dataset & Venue standard error & Int effect est & Int se & Int var & Resid Var \\
      \midrule
      Old Finals & 0.0439 & $-1.901$ & 0.0251 & 0.0019 & 0.0109 \\
      All Finals & 0.0504 & $-1.914$ & 0.0280 & 0.0025 & 0.0112 \\
      Old Pooled & 0.0495 & $-1.853$ & 0.0305 & 0.0024 & 0.0243 \\
      All Pooled & 0.0554 & $-1.869$ & 0.0348 & 0.0031 & 0.0235 \\
      \bottomrule
  \end{tabular}
  \label{tab:Gamma_parameters}
\end{table}

@@

\begin{table}
  \centering
  \caption{Compares the various effect models by dataset, all probabilities are
  for observing a reaction time less than 0.1 seconds. In general the model with
  the highest probability (least extreme reaction time) is the Year Effect model.}
  \begin{tabular}{c c c c} 
   \toprule
   Data Set & Year Effect & Batch Effect & Both Effects \\ 
   \midrule
   Old Pooled & $5.73\cdot10^{-3}$ & $5.21\cdot10^{-3}$ & $4.15\cdot10^{-3}$ \\ 
   All Pooled & $6.84\cdot10^{-3}$ & $6.64\cdot10^{-3}$ & $5.15\cdot10^{-3}$ \\
   \bottomrule
  \end{tabular}
  \label{tab:EffectComparison}
\end{table}

Using the gamma function with link set to log, the probability of the reaction 
time being less than 0.1 seconds can be calculated to see if Allen's 
disqualification was an anomaly as described in Section~\ref{sec:glmm}.
Calculations were performed in R using the methods described in 
section~\ref{sec:glmm} with the use of the \citet{lme4} package.

At a size of $n=10,000,000$, \eds{use different letter as $n$ was already used 
for number of athletes, $B?$} \of{Could I just say "At a size of 10,000,000"? 
Otherwise I am fine with using "B" instead of "n"} \eds{we may have lost the 
rest of this sentence?}

The calculated probabilities of a reaction time 
being below 0.1 seconds for the various different data sets are shown in 
Table~\ref{tab:Sim_probability}. Prior to the 2022 World Championships, there 
was a 0.0439\% chance (based on the
simulated results) of observing a reaction time below 0.01 seconds in the men's
finals.
That means that we would expect a time below the threshold to occur
once every 2,277 starts.  Considering there are roughly 8 men's finals reaction
times every 2 years (the World Championships are biennial), every 569 years there
should be a time as extreme as Allen's. This seems extremely unlikely, and thus
it seems like there may be an additional explanation for the fast reaction times.

When the 2022 data is included in the analysis, it is interesting that the 
probability of observing a reaction time below 0.1 seconds increases drastically 
for the finals data and increases by a smaller margin for the pooled data.  By
including 2022 in the data, the chances of observing an extreme time are 
increased, suggesting that 2022 is a high outlier for reaction times compared to
 other years.

Table~\ref{tab:Sim_probability} also shows that changing the reaction time 
barrier from 0.1 seconds to 0.08 seconds drastically reduces the chances of 
observing a reaction time that would break the barrier.  For the dataset of all 
finals data, the probability
associated with observing a reaction time below 0.08 seconds was $6\cdot10^{-7}$.
If IAAF \eds{or World Athletics?} were to change the reaction time the way that 
\citet{komi2009iaaf}
suggests, the probability would be so much lower that athletes would likely not
have a case to make when arguing against a disqualification.

%Discussion should show that random effect
%and standard error increase and that 2022 is special.  Now talk about Seiko
%timing equipment

@@

It is worth noting that in the semifinals of the 2022 World Track and
Field Championships, Allen's reaction time was 0.101 which is only 0.001 above
the legal limit. What this suggests is that Allen may have strong reflexes
and be able to react extremely well to the sound of the gun. It seems unlikely 
that Allen would be able to correctly predict the gun with such precision two 
times in a row, which is the exact reason that the 0.1 second reaction time 
disqualification barrier was imposed in the first place. Thus having a quick 
reaction time repeatedly may indicate skill rather than luck. Devon Allen is
an exceptional athlete, as evidenced by his football career outside of running.  
\citet{komi2009iaaf}'s paper showed that athletes are
capable of reacting in less than 0.1 seconds, and if this is the case, it is not
unreasonable to conclude that an outstanding athlete like Allen is capable of
reacting below the 0.1 second barrier.

The results from the GLMM are described in section~\ref{subsec:Results_GLMM} and
show that there

Devon Allen's disqualification at the World Track and Field Championships was
possibly the result of both a faulty timing equipment and Devon Allen
reacting extremely quickly to the start.  Our analysis showed that
the 2022 World Track and Field Championships were a low outlier compared to
every other year in terms of average reaction time.  Additionally, it was shown
through a gamma mixed effects model that there are significant year and race
effects \eds{remove race effects?}\of{I think were going to go with the two
effect model so no.} that greatly impact reaction time.

>>>>

\jy{Use double blank lines to separate paragraphs in tex source. Acknowledge if you copy.}
\of{Resolved}

@@

The dataset of the four United States athletes 
is too small to perform any meaningful statistical analysis, but
the data was expanded to include reaction times for athletes who competed at 
national competitions (many of which were held from May-July of 2022). Once the 
data was collected, each athlete was designated as a cluster and the stage of
competition (national or international) was specified for each competition 
within the cluster (athlete).  We chose to include preliminary rounds in this
section as well as in section~\ref{sec:data2019} in order to increase the size of our
data.  Without data from the preliminary rounds, the number of observations
dropped from 65 to 26 and the number of athletes included in the analysis
decreased from 15 to 9.

@@

A summary of the two effect model is shown in table~\ref{tab:Gamma_parameters}
for the pooled datasets and shows the standard error of the year
effect increases for both the men's final and men's pooled data sets when
including 2022. We choose to only look at the pooled datasets, as R displayed
warnings when creating the finals only models
\of{Is this section okay? Please let me know what I should rework}

\begin{table}
  \centering
  \caption{Summary statistics for the two effect GLMM on various data sets. SE
  is short for standard error.} 
  \begin{tabular}{c c c c c}
      \toprule
      Dataset & Batch SE & Year SE & Intercept & Residual SE \\
      \midrule
      Old Pooled & 0.060 & 0.034 & -1.865 & 0.139 \\
      All Pooled  & 0.059 & 0.041 & -1.880 & 0.014 \\
      \bottomrule
  \end{tabular}
  \label{tab:Gamma_parameters}
\end{table}

Two points of this section are to determine how extreme a reaction time below
0.1 seconds is, and to determine how the exclusion of 2022 could alter results.
Using the gamma function with link set to log, the probability of the reaction 
time being less than 0.1 seconds can be calculated in R using the methods 
described in section~\ref{sec:glmm}. A simulation size of $10,000,000$ was 
chosen to try and estimate the probability of observing a reaction time below 
two thresholds: 0.1 and 0.08. When looked at the All Pooled datset, a probability
of $5.15\cdot10^{-3}$ corresponds to roughly 1 in every 195 starts
being below 0.1 seconds.  When we exclude 2022, the probability of observing a
reaction time below 0.1 seconds decreased to $4.15\cdot10^{-3}$, or 1 in every
240 starts.
\of{I feel like with the two effect model the probability change is not as
different as it was previously}

Table~\ref{tab:Sim_probability} shows that changing the reaction time 
barrier from 0.1 seconds to 0.08 seconds drastically reduces the chances of 
observing a reaction time that would break the barrier. If World Athletics were
to change the reaction time the way that \citet{komi2009iaaf} suggests, the 
probability would be so much lower that athletes would likely not
have a case to make when arguing against a disqualification.

@@

\of{Should I report table~\ref{tab:EffectComparison}}
\begin{table}
  \centering
  \caption{Compares the various effect models by dataset, all probabilities are
  for observing a reaction time less than 0.1 seconds. In general the model with
  the highest probability (least extreme reaction time) is the Year Effect model.}
  \begin{tabular}{c c c c} 
   \toprule
   Data Set & Year Effect & Batch Effect & Both Effects \\ 
   \midrule
   Old Pooled & $5.73\cdot10^{-3}$ & $5.21\cdot10^{-3}$ & $4.15\cdot10^{-3}$ \\ 
   All Pooled & $6.84\cdot10^{-3}$ & $6.64\cdot10^{-3}$ & $5.15\cdot10^{-3}$ \\
   \bottomrule
  \end{tabular}
  \label{tab:EffectComparison}
\end{table}

@@
